---
title: "Data Wrangling with R"
author: "Alexander Alonso"
date: today
format: 
    html:
        toc: true
        toc-depth: 3
        toc-location: right-body
        embed-resources: true
execute: 
  freeze: auto
---

::: callout-note
**Note**: This codebook will follow the structure and content of the published codebook for this project: <https://jmjung.quarto.pub/m05-1-principles-data-wrangling-with-tidyverse-in-r/>
:::

# Overview

## Learning Outcomes

> By the end of this module, you will be able to:

Describe the concept of Data Wrangling.

Describe how Tibbles are different from data frames

Explain how to convert wide or long data to “Tidy” data

Explain how to merge relational data sets using join functions. (Next module)

Explain how to use grouped mutates and filter together.

Be familiar with major dplyr functions for transforming data.

Create a new variable with mutate() and case_when().

Use the pipe operator to shape the data to prepare for analysis and visualization

## The textbook chapters to cover

Ch3: Data Transfomation Ch5: Data Tidying Ch13: Numbers

# Introductin to Data Wrangling

## Loading the packages

-   to add a code chunk, use the keyboard shortcut `Ctril + Alt + I (Windows) or`Cmd + Options + I (Mac).

```{r}
library(tidyverse)
```

# What is Tidyverse?

The tidyverse is a collection of R packages that share a common design philosophy and are designed to work together seamlessly. The tidyverse includes packages for data manipulation, visualization, and modeling, among other tasks. Some of the core packages in the tidyverse include:

-   ggplot2: for data visualization

-   dplyr: for data manipulation

-   tidyr: for data tidying

-   readr: for data import

-   purrr: for functional programming

-   tibble: for data frames

-   stringr: for string manipulation

-   forcats: for working with categorical variables

## Inital Data Preparation and Exploration

```{r}
library(nycflights13)
data()
```

# Data Wranling with dplyr

Note: use 'shift + alt + I' for multi-cursor activation

-   airlines Airline names.

-   airports Airport metadata

-   flights Flights data

-   planes Plane metadata.

-   weather Hourly weather data

```{r}
library(dplyr)
library(ggplot2)
library(forcats)

flights <- flights

flights |>
  count(year, month) |>
  arrange(desc(n)) |>
  mutate(month = as.factor(month)) |>
  mutate(month, fct_reorder(month, n)) |>
  ggplot(aes(month, n, fill = year)) + 
  geom_col(fill = "#c6891f", show.legend = FALSE) +
  coord_flip()+
  labs(x = "Month",
      y = "# of Flights", 
      title = "# of Flights by Month during Year 2013 at New York"
  )
```